An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-05-30 full_paper_v2.Rmd
pre methods update

A challenge when adopting a complex systems approach is using an appropriate analysis technique that is able to capture the uncertainty and complexity of the relationships between different variables. One technique that provides a solution to this challenge is Bayesian network (BN) modelling. BN's allow the construction of graphical probabilistic models using the underlying structure that connects different variables (nodes in the network)  [@Scutari2014]. The learned structure can be used for inference by obtaining the posterior probabilities of a particular node for a given query (e.g., if the value of Node A is x and the value of Node B is y, what is the probability Node C being value z?). Furthermore, BN's do not distinguish between dependent and independent variables as they are a form of unsupervised learning, which is a strength over regression or structure equation models when the underlying relationship in the network may not be known [@Olmedilla2018].

@@

Participants self-reported any injuries they sustained at each data collection during the study period.
An injury was defined as any sports related medical problem causing the athlete to miss or modify their usual training routine. 
Minor scrapes and bruises that may require certain modifications (e.g., strapping or taping) but did not limit continued participation were not considered injuries [cf. @Appaneal2009].
Injury status (did / did not sustain an injury) served as the main outcome measure.

@@

At the start of the UK academic year (September 2016 and 2017), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. 
With the coaches' permission, the lead researcher attended training sessions to inform the athletes about the overall purpose of the study and the requirements of participation.
To be eligible for the study athletes had to be injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) and training a minimum of five hours per week. 
Athletes who met the criteria and volunteered to take part in the study were invited to attend scheduled testing sessions.

A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive twelve-month periods between September 2016 and September 2018. 
Each participant was asked to attend a total of four data collections over a twelve-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). 
Participants provided informed consent before data collection commenced.

@@

@@

To prepare the data for the BN, missing values in the dataset were first imputed. 
Out of the 668 total measurements across all time points in the current study, there were `r n_perc(missingtot$stiffness == 0)` missing muscle stiffness measurements and `r n_perc(missingtot$rmssd == 0)` missing heart rate recordings. The missing data were due to technical faults in the data collection equipment and were considered to be missing completely at random. 
A missing rate of 15-20% has been reported to be common in psychological studies, and several techniques are available to handle missing values [@Enders2003; @Lang2014].
In the current study, the _caret_ package [@Kuhn2008] was used to impute the missing values. 
A bagged tree model using all of the non-missing data was first generated and then used to predict each missing value in the dataset. 
The bagged tree method is a reliable and accurate method for imputing missing values in data and is superior to other commonly used methods such a median imputation [@Kuhn2008].

Preliminary correlation analysis of the data revealed no strong linear dependencies in the data, and therefore the data were binarised to create "Low" and "High" levels within each variable.
Binarising variables is a common procedure for BN modelling and can help improve the fit of the model by reducing the number of possible states for each node [@Beuzen2018].
All variables were approximately normally distributed; therefore, the median of each variable was used as the cut-off point to create the "Low" and "High" levels.

For the LESCA, a cumulative total of the current, and previous time points was calculated at each time point to account for the potential continuing effect of the life events experienced by athletes over time. 
Given the limited support for a relationship between positive life events and injury [@Williams2007], only negative and total life events were included in the network.
Cumulative negative and cumulative total life event scores were first log scaled and then binarised based on the median at each time point (nlelg and tlelg respectively).
In addition to the log scaled cumulative values, an untransformed NLE score from the first time point was included as an additional variable based on previous literature that indicates this variable should have a strong relationship with injury outcome [@Ivarsson2017].

@@

Prior to the final network structure presented in the results section, several structures were investigated.
Networks were learned using a Tabu search algorithm [@Norvig2009] and BIC was used to compare different models. 
A higher BIC value indicates the structure of a DAG is a better fit to the observed data [@Scutari2014]. 
BIC values for each combination of variables of interest are reported as the combination of variables with the highest BIC value, followed by the relative scores of the other variables in the model.

@@

\begin{table}[H]
\begin{adjustwidth}{-2.25in}{0in}
\caption{\label{tab:table3}Variables included in the final network structure.}
\centering
\begin{tabular}{l|l|l|l}
\hline
{\bf Variable} & {\bf Definition} & {\bf State} 1 & {\bf State 2}\\
\hline
clevel & Current competitive level & Club\_university\_county & National\_international\\
\hline
gender & Gender of the participant & Female & Male\\
\hline
hours & Number of hours spent training per week & 0-9 (Low) & >9-35 (High)\\
\hline
ind\_team & Participate in an individual or team based sport & Individual & Team\\
\hline
pi & Sustained previous injury & No Injury & Injury\\
\hline
nlebase & Untransformed NLE at TP 1 & 0-13 (Low) & >13-93 (High)\\
\hline
FFFS & Fight-Flight-Freeze System & 8-16 (Low) & >16-30 (High)\\
\hline
BIS & Behavioural Inhibition System & 17-38 (Low) & >38-68 (High)\\
\hline
RI & Reward Interest & 4-10 (Low) & >10-16 (High)\\
\hline
stiffness & Sum of all stiffness locations & 1543-2330 (Low) & >2330-4518 (High)\\
\hline
rmssd & Root mean squared difference of successive RR intervals & 2.03-4.02 (Low) & >4.02-5.94 (High)\\
\hline
balance & Total balance score & 5-15 (Low) & >15-46 (High)\\
\hline
nlelg\_1 & Log NLE at TP 1 & 0-2.64 (Low) & >2.64-4.54 (High)\\
\hline
nlelg\_2 & Log NLE at TP 2 & 0-3.04 (Low) & >3.04-5.19 (High)\\
\hline
nlelg\_3 & Log NLE at TP 3 & 0-3.18 (Low) & >3.18-4.79 (High)\\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}

Preliminary network structures also revealed strong dependencies between the same variables at subsequent time points. 
For example, the probability that stiffness_1 and stiffness_2 were both "High", or both "Low" was approximately 80%. 
Including the arcs between the same variables from X_1 $\rightarrow$ X_2 did not provide any theoretically meaningful information to the network structure as the majority of participants would be either be in a "Low" or "High" state for each pair of variables in the network. 
To more appropriately assess changes _within_ variables over time, a second BN was investigated by modelling the differences between variables at different time points.
The use of differential equations to model changes in variables over time is a common procedure in BN analysis when there are repeated measurements in the data [@Scutari2017].
To obtain the structure, variables suffixed with _1 were subtracted from variables suffixed with _2 to calculate the difference between variables measured at time points T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3. 
Independent variables were then standardized to allow relative changes between variables to be compared. 
The "injured" variable was also modified to represent whether a participant had sustained an injury at any point over the duration of the study or were healthy for the duration of the study.
The result was a network that explicitly modelled the _amount_ of change within variables between time points, as opposed to the first network that would only have captured changes when the median threshold was crossed from "Low" to "High". 
Identical blacklists to the first network were used for arcs between independent and explanatory variables. 
The nlebase variable was also dropped from the list of explanatory variables to allow the _changes_ in negative life events to be the only life event variable in the network.

>>>>

A challenge when adopting a complex systems approach is using an appropriate analysis technique that is able to capture the uncertainty and complexity of the relationships between different variables. One technique that provides a solution to this challenge is Bayesian network (BN) modelling. BN's allow the construction of graphical probabilistic models using the underlying structure that connects different variables [@Scutari2014]. The learned structure can be used for inference by obtaining the posterior probabilities of a particular variable for a given query (e.g., if the value of variable A is x and the value of variable B is y, what is the probability variable C being value z?). Furthermore, BN's do not distinguish between dependent and independent variables as they are a form of unsupervised learning, which is a strength over regression or structure equation models when the underlying relationship in the network may not be known [@Olmedilla2018].

@@

Participants self-reported any injuries they sustained at each data collection during the study period.
An injury was defined as any sports related medical problem causing the athlete to miss or modify their usual training routine for at least 24 hours [@Fuller2006; @Fuller2007b; @Timpka2014]. 
Minor scrapes and bruises that may require certain modifications (e.g., strapping or taping) but did not limit continued participation were not considered injuries [cf. @Appaneal2009].
Injury status (did / did not sustain an injury) served as the main outcome measure.

@@

At the start of the UK academic year (September 2016 and 2017), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. 
With the coaches' permission, the lead researcher attended training sessions to inform the athletes about the overall purpose of the study and the requirements of participation.
To be eligible for the study athletes had to be injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) and training a minimum of five hours per week. 
Athletes who met the criteria and volunteered to take part in the study were invited to attend scheduled testing sessions.
A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive twelve-month periods between September 2016 and September 2018. 
Each participant was asked to attend a total of 3 data collections over a twelve-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). 
Participants provided informed consent before data collection commenced.

@@

In order to use discrete networks, continuous variables must first be split into categorical levels.
When there are a large number of variables in the network, limiting the number of levels has the benefit of producing a network that is more parsimonious in terms of parameters. For example, a network with 10 variables each with two levels has 100 (10^2) possible parameter combinations, however the same network with three levels has 1000 (10^3) possible  parameter combinations, the latter being significantly more computationally expensive. Using a larger number of splits in the data also comes at a cost of reducing the statistical power in detecting probabilistic associations, and reduces the precision of parameter estimates for the probabilistic associations that are detected because it reduces the sample-size-to-parameters ratio [@Scutari2014]. Typically, no more than three levels have been used in Bayesian networks in the sports injury literature [@Olmedilla2018; @Ruiz-Perez2019]

@@

Of the 351 participants that were initially recruited for the study, 94 only completed the first time point, and therefore had to be removed from the study as no injury information was obtained for these participants following the first time point. 
To prepare the data for the BN, missing values in the dataset were first imputed. 
Out of the 650 total measurements across all time points in the current study, there were `r n_perc(missingtot$stiffness == 0)` missing muscle stiffness measurements and `r n_perc(missingtot$rmssd == 0)` missing heart rate recordings. The missing data were due to technical faults in the data collection equipment and were considered to be missing completely at random. 
A missing rate of 15-20% has been reported to be common in psychological studies, and several techniques are available to handle missing values [@Enders2003; @Lang2014].
In the current study, the _caret_ package [@Kuhn2008] was used to impute the missing values. 
A bagged tree model using all of the non-missing data was first generated and then used to predict each missing value in the dataset. 
The bagged tree method is a reliable and accurate method for imputing missing values in data and is superior to other commonly used methods such a median imputation [@Kuhn2008].

A median split technique was used to discretise the data used in the network into "Low" and "High" levels.
All variables apart from negative and total life events were approximately normally distributed (based on visual inspection, see supplementary table ...) and required no further transformation prior to the median split. 
For the LESCA questionnaire data, a cumulative total of the current, and previous time points was calculated at each time point to account for the potential continuing effect of the life events experienced by athletes over time.
Given the limited support for a relationship between positive life events and injury [@Williams2007], only negative and total life events were included in the network.
Cumulative negative, and cumulative total life event scores at each time point were first log scaled so distributions were approximately normal, and then binarised using the median at each time point (nlelg and tlelg respectively).
In addition to the log scaled cumulative values, an untransformed NLE score from the first time point was included as an additional variable based on previous literature that indicates this variable should have a strong relationship with injury outcome [@Ivarsson2017].

@@

Prior to the final network structure presented in the results section, several structures with different combinations of variables were explred.
Networks were learned using a Tabu search algorithm [@Norvig2009] and BIC was used to compare different models. 
A higher BIC value indicates the structure of a DAG is a better fit to the observed data [@Scutari2014]. 
BIC values for each combination of variables of interest are reported as the combination of variables with the highest BIC value, followed by the relative scores of the other variables in the model.

@@

```{r finalvars, results='asis'}
library(knitr)
library(kableExtra)
varcutoffs %>% 
  filter(!var %in% c("sdnn", "RR", "I", "GDP", "tlelg", "tlelg_1", "tlelg_2", "tlelg_3", "bal_asym")
) %>% 
  mutate(var = ifelse(var == "rmssd", "HRV", var)) %>% 
  kable("latex",
       booktabs = T,
       align = c("llcc"),
       caption = "Variables included in the final Bayesian network structure.",
       col.names = c("Variable", "Definition", "State 1", "State 2")) %>%
  kable_styling(latex_options = "HOLD_position") %>% 
  column_spec(2, width = c("5.2cm")) 
```

Preliminary network structures also revealed strong dependencies between the same variables at subsequent time points. 
For example, the probability that stiffness_1 and stiffness_2 were both "High", or both "Low" was approximately 80%. 
Including the arcs between the same variables from X_1 $\rightarrow$ X_2 did not provide any theoretically meaningful information to the network structure as the majority of participants would be either be in a "Low" or "High" state for each pair of variables in the network. 
To more appropriately assess changes _within_ variables over time, a second BN was investigated by modelling the differences between variables at different time points.
The use of differential equations to model changes in variables over time is a common procedure in BN analysis when there are repeated measurements in the data [@Scutari2017].
To obtain the structure, continuous variables suffixed with _1 were subtracted from variables suffixed with _2 to calculate the difference between variables measured at time points T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3. 
Independent variables were then standardized to allow relative changes between variables to be compared. 
The "injured" variable was also modified to represent whether a participant had sustained an injury at any point over the duration of the study or were healthy for the duration of the study.
The result was a network that explicitly modelled the _amount_ of change within variables between time points, as opposed to the first network that would only have captured changes when the median threshold was crossed from "Low" to "High". 
Identical blacklists to the first network were used for arcs between independent and explanatory variables. 
The nlebase variable was also dropped from the list of explanatory variables to allow the _changes_ in negative life events to be the only life event variable in the network.
